Kindly Bent to Free Us

### 2019-11-12 kindly/intro.tex
Swap "mutable/immutable borrow" for "exclusive/shared borrow".

Our work provides a solution to programmers in this situation. We
propose an extended type system for ML-like languages that comes with
linear and affine types, mutable and immutable borrows, but all that
integrated with full principal type inference, garbage collected data, and
automatic placement of borrowing regions.

The most closely related contenders in this design space are Linear Haskell
\cite{DBLP:journals/pacmpl/BernardyBNJS18}, henceforth called LH,
Quill \cite{DBLP:conf/icfp/Morris16}, and ALMS \cite{DBLP:conf/popl/TovP11}.
Compared to LH, the goals and means are
similar as both systems permit abstraction over the number of uses of
values and retain type inference, but the details are quite different.
\begin{enumerate}
\item LH's multiplicities are either linear or unrestricted whereas
  we also permit affine values.
\item Our system supports mutable and immutable borrows, hence the
  need for affinity.
\item Our multiplicities are directly attached to the type of the
  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
  denotes the type of a \emph{single-use function} that can be called
  just once, whereas the multiplicities in LH choose
  between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
  latter is a function that promises to \emph{use its argument exactly
    once}.
\item Similar to Quill, we make use of multiplicity contraints. We
  rely on kind subsumption to simplify the inferred types.
\end{enumerate}
See \cref{sec:related-work} for further in-depth discussion of these
and other related works.
\subsection{First examples}
\label{sec:first-example}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.45\linewidth}
    \lstinputlisting{code/writefiles.affe}
    \vspace{-15pt}
    \caption{File API}
    \label{fig:writing-files-api}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\linewidth}
\begin{lstlisting}
let main () =
  let h = File.fopen "foo" in
  File.write &!h "Hello ";
  File.write &!h "world!";
  File.close h
\end{lstlisting}
    \vspace{-10pt}
    \caption{File example}
    \label{fig:writing-files-example}
  \end{subfigure}
  \vspace{-10pt}
  \caption{Writing files}
  \label{fig:writing-fiules}
\end{figure}

As a first, well-known example we consider a simplified API for
writing files shown in \cref{fig:writing-files-api}.  It introduces a
linear abstract type \lstinline/File.t/. A call like
\lstinline/fopen "foo"/ returns a linear handle to a newly created
file, which \emph{must} be released later on through \lstinline/close/
as shown in \cref{fig:writing-files-example}. Failing to do so is a
static type error.  To write to the file, we must take a (mutable)
borrow \lstinline/&!h/ of the handle and pass it to the
\lstinline/write/ function. Mutable borrows are classified as affine:
they must not be duplicated, but there is no requirement to use
them. This affinity shows up in the annotation \lstinline/-{aff}>/ of the second arrow in
the type of \lstinline/write/: a partial application like
\lstinline/write &!h/ captures the affine borrow and hence the
resulting function must also be handled in an affine way. That is, it
would be an error to write
\begin{lstlisting}
let w = File.write &!h in w "Hello "; w "world!" (*error*)
\end{lstlisting}
as we would be using the affine closure twice. The remaining arrows in
the API are unrestricted and we write \lstinline/->/ instead of the
explicitly annotated \lstinline/-{un}>/.

>>>>

Our work provides a solution to programmers in this situation. We
propose an extended type system for ML-like languages that comes with
linear and affine types, exclusive and shared borrows, but all that
integrated with full principal type inference, garbage collected data, and
automatic placement of borrowing regions.

The most closely related contenders in this design space are Linear Haskell
\cite{DBLP:journals/pacmpl/BernardyBNJS18}, henceforth called LH,
Quill \cite{DBLP:conf/icfp/Morris16}, and ALMS \cite{DBLP:conf/popl/TovP11}.
Compared to LH, the goals and means are
similar as both systems permit abstraction over the number of uses of
values and retain type inference, but the details are quite different.
\begin{enumerate}
\item LH's multiplicities are either linear or unrestricted whereas
  we also permit affine values.
\item Our system supports exclusive and shared borrows, hence the
  need for affinity.
\item Our multiplicities are directly attached to the type of the
  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
  denotes the type of a \emph{single-use function} that can be called
  just once, whereas the multiplicities in LH choose
  between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
  latter is a function that promises to \emph{use its argument exactly
    once}.
\item Similar to Quill, we make use of multiplicity contraints. We
  rely on kind subsumption to simplify the inferred types.
\end{enumerate}
See \cref{sec:related-work} for further in-depth discussion of these
and other related works.
\subsection{First examples}
\label{sec:first-example}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.45\linewidth}
    \lstinputlisting{code/writefiles.affe}
    \vspace{-15pt}
    \caption{File API}
    \label{fig:writing-files-api}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\linewidth}
\begin{lstlisting}
let main () =
  let h = File.fopen "foo" in
  File.write &!h "Hello ";
  File.write &!h "world!";
  File.close h
\end{lstlisting}
    \vspace{-10pt}
    \caption{File example}
    \label{fig:writing-files-example}
  \end{subfigure}
  \vspace{-10pt}
  \caption{Writing files}
  \label{fig:writing-fiules}
\end{figure}

As a first, well-known example we consider a simplified API for
writing files shown in \cref{fig:writing-files-api}.  It introduces a
linear abstract type \lstinline/File.t/. A call like
\lstinline/fopen "foo"/ returns a linear handle to a newly created
file, which \emph{must} be released later on through \lstinline/close/
as shown in \cref{fig:writing-files-example}. Failing to do so is a
static type error.  To write to the file, we must take a (exclusive)
borrow \lstinline/&!h/ of the handle and pass it to the
\lstinline/write/ function. Exclusive borrows are classified as affine:
they must not be duplicated, but there is no requirement to use
them. This affinity shows up in the annotation \lstinline/-{aff}>/ of the second arrow in
the type of \lstinline/write/: a partial application like
\lstinline/write &!h/ captures the affine borrow and hence the
resulting function must also be handled in an affine way. That is, it
would be an error to write
\begin{lstlisting}
let w = File.write &!h in w "Hello "; w "world!" (*error*)
\end{lstlisting}
as we would be using the affine closure twice. The remaining arrows in
the API are unrestricted and we write \lstinline/->/ instead of the
explicitly annotated \lstinline/-{un}>/.

### 2019-11-12 kindly/lang.tex
Swap "mutable/immutable borrow" for "exclusive/shared borrow".

We start from the bottom.
A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is mutable
($\MBORROW$) or immutable ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.
%
A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
%
A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).

@@

\cref{sdtyp:split} shows the action of the splitting rules on single
bindings. If $x$'s type is unrestricted, then
the {\sc Both} rule indicates that we can duplicate it.
Similarly, unrestricted borrows can be duplicated according to rule
{\sc Borrow}.
The {\sc Left} and {\sc Right} rules are always applicable and move a binding
either to the left or right environment.
The rules {\sc Susp} and {\sc SuspB} split off suspended bindings to
the left while conserving access to the binding on the right.
A suspended binding can later be turned
into a borrow inside a region. Splitting of suspended bindings is
asymmetric. It must follow the order of execution from left to right,
which means that a resource can be used first as a borrow on the left
and then later as a full resource on the right. The {\sc Susp} rule
works with a full resource while the {\sc SuspB}
rule splits off an immutable suspended binding from a borrow.
%
Splitting applies whenever an
expression has multiple subexpressions:  pairs, function applications,
and let bindings. In the
expression
$\letin{a}{\text{create}\ 10\ x}
{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
the rule {\sc Susp} splits off a borrow from  the resource
\lstinline/a/ to use it in the left component of the pair.
As usual, a borrow cannot be active in the same scope as its resource.
The \emph{region} around its use ensures that the borrow in the left component does not
escape, which brings us to the next topic.

>>>>

We start from the bottom.
A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive
($\MBORROW$) or shared ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.
%
A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
%
A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).

@@

\cref{sdtyp:split} shows the action of the splitting rules on single
bindings. If $x$'s type is unrestricted, then
the {\sc Both} rule indicates that we can duplicate it.
Similarly, unrestricted borrows can be duplicated according to rule
{\sc Borrow}.
The {\sc Left} and {\sc Right} rules are always applicable and move a binding
either to the left or right environment.
The rules {\sc Susp} and {\sc SuspB} split off suspended bindings to
the left while conserving access to the binding on the right.
A suspended binding can later be turned
into a borrow inside a region. Splitting of suspended bindings is
asymmetric. It must follow the order of execution from left to right,
which means that a resource can be used first as a borrow on the left
and then later as a full resource on the right. The {\sc Susp} rule
works with a full resource while the {\sc SuspB}
rule splits off an shared suspended binding from a borrow.
%
Splitting applies whenever an
expression has multiple subexpressions:  pairs, function applications,
and let bindings. In the
expression
$\letin{a}{\text{create}\ 10\ x}
{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
the rule {\sc Susp} splits off a borrow from  the resource
\lstinline/a/ to use it in the left component of the pair.
As usual, a borrow cannot be active in the same scope as its resource.
The \emph{region} around its use ensures that the borrow in the left component does not
escape, which brings us to the next topic.

### 2019-11-12 kindly/motivation.tex
Swap "mutable/immutable borrow" for "exclusive/shared borrow".

To manage the different kinds of accessing the array we distinguish between constructors, destructors,
observers, and mutators.
Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
array. As with channels in \cref{sec:session-linearity},
\lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
During the lifetime of the array resource \lstinline/a/, we can split
off \emph{immutable borrows} \lstinline/&a/ that provide a read-only
view or \emph{mutable borrows} \lstinline/&!a/ for read-write views.
Observer functions such as \lstinline/length/ and \lstinline/get/ expect an immutable borrow
argument, mutator functions such a \lstinline/set/ expect
a mutable borrow.

Each borrow is tied to a region whose lifetime is properly
contained in the lifetime of the resource.
In a region, we can split off as many immutable borrows as we like
from a resource, but we can take only one mutable borrow. In a
subsidiary region, we can take further immutable borrows from any
borrow or we can take a mutable borrow from a mutable borrow from a
enclosing region. Borrows are confined to their regions. Inside the region,
immutable borrows are unrestricted (\lstinline/un/) whereas mutable
borrows are affine (\lstinline/aff/).

The code in \cref{ex:array} demonstrates a use of the API to create an
array of Fibonacci numbers. After creation of the array, we can see
that the presence of a borrow prevents access to the ``raw''
resource inside the borrow's region. In \cref{ex:array:get} of the
example, there are two immutable borrows in the same
expression, which forms a region by itself. These borrows are split
off the mutable borrow used in \cref{ex:array:set}, which belongs to
the next enclosing region.
Finally, the code returns the whole array in \cref{ex:array:return},
which is possible because none of the borrows are in scope anymore.

@@

\cref{sig:hybarray} contains an API \lstinline/HybArray/
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
The in-place mutation function \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ works on a mutable borrow  whereas the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an immutable borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
\lstinline/mapi : (int -> &'a -> 'b) -> &('a t) -> 'b t /
is a simple variation on \lstinline/Array.map/ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the \lstinline/Matrix/ type uses the same API as \lstinline/HybArray/
but with two indices.
Each cell contains an integer set that represent admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choice point,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the \lstinline/propagate/ function.
The \lstinline/propagate/ function uses direct mutation through a
mutable borrow of the matrix as it need not preserve the previous
version of the board.
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square, which are all very similar to function
\lstinline/propagate_lines/ (\cref{line:propline}).

The typing ensures that the mutations do not compromise the state at
the choice point, because they operate on a new state \lstinline/new_g/ created for one
particular branch of the choice.
In particular,
since the \lstinline/set/ function only requires an unrestricted immutable borrow,
\lstinline/try_solution/ can captures the borrow \lstinline/&g/
while staying unrestricted.
This also make \lstinline/try_solution/ itself bound to the same
region as \lstinline/&g/. In this example, our region inference
will place the region exactly at the \lstinline/begin//\lstinline/end/
scope.
This allows us to use \lstinline/try_solution/ in
the iteration loop on \cref{line:iter}.
Since \lstinline/&g/ is unrestricted, we could even use a parallel iteration
instead of a sequential one.
As \lstinline/g/ is linear we must free it before returning,
in \cref{line:free:g}, outside of the region.

@@

\cref{intf:pool} contains the interface of the \lstinline/Pool/ module.
A pool is parameterized by its content. The kind of the pool
depends on the content: linear content implies
a linear pool while unrestricted content yields an unrestricted pool.
The functions \lstinline/Pool.create/ and \lstinline/Pool.consume/
build/destroy a pool given creators/destructors for the elements
of the pool.
The function \lstinline/Pool.use/ is the workhorse of the API, which
borrows a resource from the pool to a callback.
It takes a shared borrow of a pool (to enable concurrent access) and a
callback function.
The callback receives a mutable borrow of an arbitrary resource from the pool.
The typing of the callback ensures
that this borrow is neither captured nor returned by the function.

This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
The signature prescribes the type \lstinline/&!(aff_r1,'a_1 )/ for the
mutable borrow of the resource with an affine kind at region nesting $r+1$. The return
type of the callback is constrained to kind \lstinline/'k_2 <= aff_r/
so that the callback certainly cannot return the borrowed argument.
In a specific use of \lstinline/Pool.use/, the index $r$ gets unified
with the current nesting level of regions so that the region for the
callback effectively gets ``inserted'' into the lexical nesting at the callsite.

>>>>

To manage the different kinds of accessing the array we distinguish between constructors, destructors,
observers, and mutators.
Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
array. As with channels in \cref{sec:session-linearity},
\lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
During the lifetime of the array resource \lstinline/a/, we can split
off \emph{shared borrows} \lstinline/&a/ that provide a read-only
view or \emph{exclusive borrows} \lstinline/&!a/ for read-write views.
Observer functions such as \lstinline/length/ and \lstinline/get/ expect an shared borrow
argument, mutator functions such a \lstinline/set/ expect
a exclusive borrow.

Each borrow is tied to a region whose lifetime is properly
contained in the lifetime of the resource.
In a region, we can split off as many shared borrows as we like
from a resource, but we can take only one exclusive borrow. In a
subsidiary region, we can take further shared borrows from any
borrow or we can take a exclusive borrow from a exclusive borrow from a
enclosing region. Borrows are confined to their regions. Inside the region,
shared borrows are unrestricted (\lstinline/un/) whereas exclusive
borrows are affine (\lstinline/aff/).

The code in \cref{ex:array} demonstrates a use of the API to create an
array of Fibonacci numbers. After creation of the array, we can see
that the presence of a borrow prevents access to the ``raw''
resource inside the borrow's region. In \cref{ex:array:get} of the
example, there are two shared borrows in the same
expression, which forms a region by itself. These borrows are split
off the exclusive borrow used in \cref{ex:array:set}, which belongs to
the next enclosing region.
Finally, the code returns the whole array in \cref{ex:array:return},
which is possible because none of the borrows are in scope anymore.

@@

\cref{sig:hybarray} contains an API \lstinline/HybArray/
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
The in-place mutation function \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ works on a exclusive borrow  whereas the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
\lstinline/mapi : (int -> &'a -> 'b) -> &('a t) -> 'b t /
is a simple variation on \lstinline/Array.map/ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the \lstinline/Matrix/ type uses the same API as \lstinline/HybArray/
but with two indices.
Each cell contains an integer set that represent admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choice point,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the \lstinline/propagate/ function.
The \lstinline/propagate/ function uses direct mutation through a
exclusive borrow of the matrix as it need not preserve the previous
version of the board.
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square, which are all very similar to function
\lstinline/propagate_lines/ (\cref{line:propline}).

The typing ensures that the mutations do not compromise the state at
the choice point, because they operate on a new state \lstinline/new_g/ created for one
particular branch of the choice.
In particular,
since the \lstinline/set/ function only requires an unrestricted shared borrow,
\lstinline/try_solution/ can captures the borrow \lstinline/&g/
while staying unrestricted.
This also make \lstinline/try_solution/ itself bound to the same
region as \lstinline/&g/. In this example, our region inference
will place the region exactly at the \lstinline/begin//\lstinline/end/
scope.
This allows us to use \lstinline/try_solution/ in
the iteration loop on \cref{line:iter}.
Since \lstinline/&g/ is unrestricted, we could even use a parallel iteration
instead of a sequential one.
As \lstinline/g/ is linear we must free it before returning,
in \cref{line:free:g}, outside of the region.

@@

\cref{intf:pool} contains the interface of the \lstinline/Pool/ module.
A pool is parameterized by its content. The kind of the pool
depends on the content: linear content implies
a linear pool while unrestricted content yields an unrestricted pool.
The functions \lstinline/Pool.create/ and \lstinline/Pool.consume/
build/destroy a pool given creators/destructors for the elements
of the pool.
The function \lstinline/Pool.use/ is the workhorse of the API, which
borrows a resource from the pool to a callback.
It takes a shared borrow of a pool (to enable concurrent access) and a
callback function.
The callback receives a exclusive borrow of an arbitrary resource from the pool.
The typing of the callback ensures
that this borrow is neither captured nor returned by the function.

This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
The signature prescribes the type \lstinline/&!(aff_r1,'a_1 )/ for the
exclusive borrow of the resource with an affine kind at region nesting $r+1$. The return
type of the callback is constrained to kind \lstinline/'k_2 <= aff_r/
so that the callback certainly cannot return the borrowed argument.
In a specific use of \lstinline/Pool.use/, the index $r$ gets unified
with the current nesting level of regions so that the region for the
callback effectively gets ``inserted'' into the lexical nesting at the callsite.

### 2019-11-12 kindly/sdtyping/extra.tex
Swap "mutable/immutable borrow" for "exclusive/shared borrow".

So far, all \lang programs have been fully annotated with regions information.
We now show how to infer these regions annotations based on
optionally-annotated programs.
First, we extend the region annotation to $\region{S}{E}$ where $S$ is
a map from variables to borrow indicator $b$. This annotation, defined below, is equivalent to nested
region annotations for each individual variable.
%
\begin{align*}
  \region{\Sone x\BORROW;S}{e} &= \region{\Sone x\BORROW}{\region{S}{e}}& \region{\emptyset}{e} &= e\\
\end{align*}
%
\begin{figure*}[!tb]
  \centering
  \input{region-annot}
  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
  \label{fig:region-annotation}
\end{figure*}
%
\Cref{fig:region-annotation} define a rewriting relation $\RannotT{e}{e'}$
which indicates that an optionally annotated term $e$ can be rewritten
as a fully annotated term $e'$.
Through the rule \textsc{Rewrite-Top}, this is defined
in term of an inductively defined relation
$\Rannot{e}{e'}{S}$ where $n$ is the current nesting and $S$ is a set of
variable that are not yet enclosed in a region.
The base cases are constants, variables and borrows.
The general idea is to start from the leaves of the syntax tree, create a
region for each borrow, and enlarge the region as much as possible.
This is implemented by a depth-first walk of the syntax
tree which collects each variable that has a corresponding borrow.
At each step, it rewrites the inner subterms,
consider which borrow must be enclosed by a region now, and
return the others for later enclosing. Binders force immediate
enclosing of the bound variables, as demonstrated in rule \textsc{Rewrite-Lam}.
For nodes with multiple children, we
use a scope merge operator to decide if regions should be placed and where.
This is shown in rule \textsc{Rewrite-Pair}.
The merge operator, written $\getBorrows{B_l}{B_r}{(S_l,S,S_r)}$, takes
the sets $B_l$ and $B_r$ returned by rewriting the subterms
and returns three sets: $S_l$ and $S_r$ indicates the variables
that should be immediately enclosed by a region on the left and right
subterms and $S$ indicates the set of the yet-to-be-enclosed variables.
As an example, the rule \textsc{AnnotRegion-MutLeft} is applied
when there is an immutable borrow and a mutable borrow. In that case, a
region is created to enclose the immutable borrow, while the mutable
borrow is left to be closed later. This is coherent with the rules
for environment splitting and suspended bindings from \cref{sdtyping}.
%
Explicitly annotated regions are handled specially through
rule \textsc{Rewrite-Region}. In that case, we assume that all inner
borrows should be enclosed immediately.

@@

\begin{figure*}[!tb]
  \input{sdtyping/sd-splitting}
  \caption{Splitting ---
    environments $\lsplit C\E{\E_l}{\E_r}$;
    inference $\bsplit C\E{\bm\E_l}{\bm\E_r}$;
    binders $\bsplit Cb{\bm b_r}{\bm b_l}$}
  \label{fig:sd-splitting}
  \input{sdtyping/sd-borrowing}
  \caption{Borrowing ---
    environments $\lregion[n]{C}{x}{\E}{\E'}$;
    inference $\bregion[\bm n]{C}{\bm x}{\E}{\bm\E'}$;
    binders $\bregion[\bm n]{C}{\bm x}{b}{\bm b'}$}
  \label{fig:sd-borrowing}
  \input{sdtyping/sd-environments}
  \caption{Rewriting constraints on environments --- $\bm{\Cleq{\E}{k}}\Crewrite  C$}
  \label{fig:contraints-environments-types}
\end{figure*}

@@

\begin{figure*}[hbtp]
  \input{sdtyping/sd-typing}
  \caption{Syntax-directed typing rules}
  \label{fig:syntax-directed-typing}
\end{figure*}

>>>>

So far, all \lang programs have been fully annotated with regions information.
We now show how to infer these regions annotations based on
optionally-annotated programs.
First, we extend the region annotation to $\region{S}{E}$ where $S$ is
a map from variables to borrow indicator $b$. This annotation, defined below, is equivalent to nested
region annotations for each individual variable.
%
\begin{align*}
  \region{\Sone x\BORROW;S}{e} &= \region{\Sone x\BORROW}{\region{S}{e}}& \region{\emptyset}{e} &= e\\
\end{align*}
%
\begin{figure*}[!tb]
  \centering
  \input{region-annot}
  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
  \label{fig:region-annotation}
\end{figure*}
%
\Cref{fig:region-annotation} define a rewriting relation $\RannotT{e}{e'}$
which indicates that an optionally annotated term $e$ can be rewritten
as a fully annotated term $e'$.
Through the rule \textsc{Rewrite-Top}, this is defined
in term of an inductively defined relation
$\Rannot{e}{e'}{S}$ where $n$ is the current nesting and $S$ is a set of
variable that are not yet enclosed in a region.
The base cases are constants, variables and borrows.
The general idea is to start from the leaves of the syntax tree, create a
region for each borrow, and enlarge the region as much as possible.
This is implemented by a depth-first walk of the syntax
tree which collects each variable that has a corresponding borrow.
At each step, it rewrites the inner subterms,
consider which borrow must be enclosed by a region now, and
return the others for later enclosing. Binders force immediate
enclosing of the bound variables, as demonstrated in rule \textsc{Rewrite-Lam}.
For nodes with multiple children, we
use a scope merge operator to decide if regions should be placed and where.
This is shown in rule \textsc{Rewrite-Pair}.
The merge operator, written $\getBorrows{B_l}{B_r}{(S_l,S,S_r)}$, takes
the sets $B_l$ and $B_r$ returned by rewriting the subterms
and returns three sets: $S_l$ and $S_r$ indicates the variables
that should be immediately enclosed by a region on the left and right
subterms and $S$ indicates the set of the yet-to-be-enclosed variables.
As an example, the rule \textsc{AnnotRegion-MutLeft} is applied
when there is an shared borrow and a exclusive borrow. In that case, a
region is created to enclose the shared borrow, while the exclusive
borrow is left to be closed later. This is coherent with the rules
for environment splitting and suspended bindings from \cref{sdtyping}.
%
Explicitly annotated regions are handled specially through
rule \textsc{Rewrite-Region}. In that case, we assume that all inner
borrows should be enclosed immediately.

@@

\begin{figure*}[!tb]
  \input{sdtyping/sd-splitting}
  \caption{Splitting ---
    environments $\lsplit C\E{\E_l}{\E_r}$;
    inference $\bsplit C\E{\bm{\E_l}}{\bm{\E_r}}$;
    binders $\bsplit Cb{\bm{b_r}}{\bm{b_l}}$}
  \label{fig:sd-splitting}
  \input{sdtyping/sd-borrowing}
  \caption{Borrowing ---
    environments $\lregion[n]{C}{x}{\E}{\E'}$;
    inference $\bregion[\bm n]{C}{\bm x}{\E}{\bm{\E'}}$;
    binders $\bregion[\bm n]{C}{\bm x}{b}{\bm{b'}}$}
  \label{fig:sd-borrowing}
  \input{sdtyping/sd-environments}
  \caption{Rewriting constraints on environments --- $\bm{\Cleq{\E}{k}}\Crewrite  C$}
  \label{fig:contraints-environments-types}
\end{figure*}

@@

\begin{figure*}[hbtp]
  \input{sdtyping/sd-typing}
  \caption{Syntax-directed typing rules --
    $\inferS{C}{\E}{e}{\tau}$}
  \label{fig:syntax-directed-typing}
\end{figure*}
